Smart Representations: {R}ationality and Evolution in a Richer Environment

### 2015-12-16 paper/graph.tex
rewrite intro

    \node[rounded corners, draw, rectangle, minimum height = 0.75cm, text width=6cm, text
    centered, below of = environment, node distance = 2cm] (situation)
    {\textbf{decision situation} $d$};

@@

    \draw[->, thick]
    ([xshift=-1.75cm]environment.center)--([xshift=-1.75cm,yshift=0.5cm]situation.center)
    node[midway, right] {environment presents random $d$};

@@

    \begin{pgfonlayer}{background}
             \node [rounded corners,
       very thick,draw=black!40,fit={($(situation.south)+(0,+30pt)$) 
                                 ($(mechanism.west)+(+5pt,0)$) 
                                 ($(action.north)+(0,-30pt)$) 
                                 ($(representation.west)+(-7pt,0)$)}] (box) {};
    \end{pgfonlayer}

>>>>

    \node[rounded corners, draw, rectangle, minimum height = 0.75cm, text width=6cm, text
    centered, below of = environment, node distance = 2cm] (situation)
    {environment produces random \\ \textbf{decision situation} $d$};

@@

@@

    \begin{pgfonlayer}{background}
             \node [rounded corners, dashed,
       very thick,draw=black!40,fit={($(situation.south)+(0,-10pt)$) 
                                     ($(mechanism.west)+(+5pt,+42pt)$) 
                                     % ($(action.north)+(0,+10pt)$) 
                                     ($(representation.west)+(-7pt,0)$)
                                   }] (box) {};
    \end{pgfonlayer}

### 2015-12-16 paper/paper.tex
rewrite intro

%========================================
% Standard Layout
%========================================

@@

\begin{quotation} \textit{This is what I aim at, because the point of philosophy is to start with something so simple as not to seem worth stating, and to end with something so paradoxical that no one will believe it.} \citep{Russ18} \end{quotation}

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modelling. However, most applications
look at evolutionary selection of behavior in a single \emph{stage game}. There are two common
simplifications in this approach. Firstly, the stage game is usually considered fixed, closed
and immutable. Secondly, what evolves are behavioral strategies for the stage game. It has
consequently been argued that, instead of focusing on a fixed, immutable stage game and the
behavioral selection for that stage game alone, we should rather study the evolutionary
competition of \emph{choice mechanisms} under selective pressure in a rich and variable
environment \citep[e.g.,][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}. 
But only little attention has been paid to modeling the evolutionary competition between choice
mechanisms in a game-theoretical framework \citep[but
see][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}.\todo{More references
  Harley 1981!!!, O Conor \dots}

If agents deal with a rich and varied environment, they face many different choice
situations. A crucial part of any choice mechanism is a \emph{subjective representation} of the
choice situation, e.g., the agent's manner of representing their preferences and her way of
forming decision relevant beliefs in the case of uncertainty. Since almost any reasonable
learning or decision making process will make use of some (subjective, agent-internal, possibly
unconscious) representation of both beliefs and the quality of choice options, the question of which such
representations are ecologically valuable (i.e., lead to high fitness) is central and has been
subject of recent interest also in economics literature \citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. Na\"ive intuition might convince us that preferences that veridically
represent the actual, true fitness measure that evolutionary selection is based upon must
surely be the best. We demonstrate that, if we take into account the different ways agents may cope with
uncertainty, this na\"ive conclusion is premature. Under certain reasonable epistemic
conditions, non-veridical preference representations can be evolutionarily beneficial.

To show this, we study the evolutionary competition of choice mechanisms, or more specifically
the competition of subjective representations of choice situations, in a variable
environment. Concretely, we look at evolutionary ``meta-games'' whose ``acts'' are choice mechanisms and whose ``payoffs'' represent the average expected
fitness that different choice mechanisms would accrue when playing
arbitrary games (from a given class, with a given occurrence probability). In this sense, a
meta-game captures (the modeller's) assumptions about the relevant statistics of the
environment in which evolutionary forces operate, while we are still able to use standard
methods from evolutionary game theory, like stability or evolutionary dynamics.

Section \ref{sec:rati--subj} overviews different accounts of rationality in decision making and relates the approach we take here to the existing literature on the topic. Section \ref{sec:basic-notions} introduces the various decision- and game-theoretic notions that will be later used in the evolutionary analysis. Sections \ref{sec:model} and \ref{sec:basic-model-1} define the evolutionary framework, and Section \ref{sec:results:-basic-model} presents the main results of the paper. Section \ref{sec:extensions} extends the evolutionary analysis by introducing more choice mechanisms in the model and by investigating the evolutionary performances of different choice mechanisms when facing arbitrary decision problems. Finally, in Section \ref{sec:conclusion} we comment on the results achieved and future perspectives.

\section{Rationality, subjectivity, beliefs \& preferences}
\label{sec:rati--subj}

@@

If beliefs and preferences are subjective, there is room for \emph{rationalization} or
\emph{redescriptionism} of observable behavior. Prospect theory is, perhaps, the most prominent
example. \citet{KahnemannTversky1979:Prospect-Theory} looked at experimental data from human
choices in suitably defined tasks and demonstrated that what appears to be violations of
rationality norms can be explained on the assumption that subjects' beliefs and preferences
deviate systematically from the objectively given parameters in the presented choice task
(e.g., by taking into account reference points and diminishing returns for both gains and
losses).  Similarly, tinkering with the notion of subjective utility, in particular by
including social preferences such as fairness, reciprocity or similar, allows us to describe as
rational empirically observed behavior in situations of social decision making, such as the
prisoners' dilemma or public goods games, that might otherwise appear irrational \citep[e.g.,][]{fehrschmidt99,charrab02}.

It could be objected that, without additional constraints, the notion of rationality is likely
to collapse, as it seems possible to deem rational almost everything that is observed, given the
freedom to adjust beliefs and preferences at will. But there are many ways in which ascriptions
of beliefs and preferences are constrained by normative considerations of rationality. To begin
with, if the DM's uncertainty is about objective chance, it can be argued that subjective
beliefs should conform (at least roughly) to objective chance. When the DM's uncertainty cannot
be anchored in an objective notion of chance, then temporal consistency in his decision making
should at least constrain his subjective beliefs, else he fall prey to exploitation by Dutch
books. Similarly, in certain situations it may make sense to postulate on normative grounds
that subjective preferences should be oriented towards tracking objective utility, e.g.,
fitness. For instance, in their classic work \myalert{(Alchian, Friedman)} argue that profit
maximization is a reasonable assumption for characterizing outcomes in a competitive market
because only firms behaving in a manner that is consistent with profit maximization will
survive in the long run.\todo{@Paolo: please add references}

An alternative to descriptive vs.~normative views on rationality is \emph{adaptationism}
\citep[e.g.,][]{HagenChater2012:Decision-Making}. \todo{more references} Adaptationism aims to
explain choice behavior by appeal to evolutionary considerations. The key idea is that decision
makers (human and other life forms) have evolved choice mechanisms which were fitness enhancing
in the relevant domains of choice making, at the relevant time of evolution. When it comes to
preferences, this is clearly related to Alchian and Friedman's argument that firms
\emph{should} optimize that which makes them ``live,'' so to speak. For biological agents, the
argument would be that subjective preferences \emph{should} stay as close as possible to whatever determines objective
fitness.

>>>>

\usetikzlibrary{positioning,arrows,calc,fit}
%========================================
% Standard Layout
%========================================

@@

% \begin{quotation} \textit{This is what I aim at, because the point of philosophy is to start with something so simple as not to seem worth stating, and to end with something so paradoxical that no one will believe it.} \citep{Russ18} \end{quotation}

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modelling at a high level of
abstraction.  Abstraction entails simplifying assumptions. Evolutionary game models heavily
simplify reality in many ways as well. Two simplifications relevant here are that the
environment is represented as a fixed, closed and immutable \emph{stage game} and that the
focus of evolutionary selection is behavior for this stage game alone. In contrast, some argue
for studying instead the evolutionary competition of general \emph{choice mechanisms} in a rich
and variable environment
\citep[e.g.,][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}. To show that
there is no tension here, this paper introduces a method for reconciling established notions of
evolutionary game theory with evolutionary selection of choice mechanisms in variable
environments \citep[see related work
by][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}.\todo{More references
  Harley 1981!!!, O Conor \dots}

If agents deal with a rich and varied environment, they face many different choice
situations. A crucial part of any choice mechanism is a \emph{subjective representation} of the
choice situation, in particular a manner of forming preferences and beliefs about a possibly
uncertain world. This paper asks: which preference and belief representations are ecologically
valuable (i.e., lead to high fitness)? The evolution of preferences has been subject of recent
interest in theoretical economics \citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. The
particular contribution of this paper is to provide a case, spelled out formally, that
questions about the evolution of preferences cannot be separated from questions about the
evolution of beliefs. As a point in case, we demonstrate that if agents have imprecise
probabilistic beliefs \citep[e.g.][]{gardsah82,levi74,walley96}, preference representations in
terms of regret that deviate from the
true, objective fitness that evolutionary selection operates on can be selected for.

% To show this, we study the evolutionary competition of choice mechanisms, or more
% specifically the competition of subjective representations of choice situations, in a
% variable environment. Concretely, we look at evolutionary ``meta-games'' whose ``acts'' are
% choice mechanisms and whose ``payoffs'' represent the average expected fitness that different
% choice mechanisms would accrue when playing arbitrary games (from a given class, with a given
% occurrence probability). In this sense, a meta-game captures (the modeller's) assumptions
% about the relevant statistics of the environment in which evolutionary forces operate, while
% we are still able to use standard methods from evolutionary game theory, like stability or
% evolutionary dynamics.

Section \ref{sec:rati--subj} reviews different accounts of rationality in decision theory and
relates the approach we take here to the existing literature on the topic. Section
\ref{sec:basic-notions} introduces relevant decision- and game-theoretic notions. Sections
\ref{sec:model} and \ref{sec:basic-model-1} define the evolutionary framework, and Section
\ref{sec:results:-basic-model} presents the main results of the paper. Section
\ref{sec:extensions} extends the evolutionary analysis by introducing more choice mechanisms in
the model and by investigating the evolutionary performances of different choice mechanisms
when facing arbitrary decision problems. Finally, in Section \ref{sec:conclusion} we reflect on
the results achieved.

\todo[inline]{rewrite overview after restructuring}

\section{Rationality, subjective beliefs \& preferences and the evolution of choice mechanisms}
\label{sec:rati--subj}

@@

If beliefs and preferences are subjective, there is room for \emph{rationalization} or
\emph{redescriptionism} of observable behavior. For example,
\citet{KahnemannTversky1979:Prospect-Theory} famously demonstrated that what appear to be
violations of rationality norms in human decision making can be explained on the assumption
that subjects' beliefs and preferences deviate systematically from the objectively given
parameters in the presented choice task. Similarly for social decision making, including social
preferences such as fairness, reciprocity or similar, allows us to describe as rational
empirically observed behavior, such as in experimental prisoners' dilemmas or public goods
games, that might otherwise appear irrational \citep[e.g.,][]{fehrschmidt99,charrab02}.

The main objection to redescriptionism is that, without additional constraints, the notion of
rationality is likely to collapse, as it seems possible to deem rational almost everything that
is observed, given the freedom to adjust beliefs and preferences at will. \emph{Normativism}
therefore emphasizes that there are many ways in which ascriptions of beliefs and preferences
are constrained by normative considerations of rationality as well. As for beliefs, where
possible, subjective beliefs should reflect objective chance. Where this is impossible,
temporal consistency of choices constrains rational subjective beliefs, else exploitation by
Dutch books is possible. As for preferences, in certain situations it may make sense to
postulate on normative grounds that subjective preferences should be oriented towards tracking
objective fitness. For instance, profit maximization seems a reasonable assumption for
characterizing outcomes in a competitive market because only firms behaving in a manner that is
consistent with profit maximization will survive in the long run \myalert{(Alchian,
  Friedman)}.\todo{@Paolo: please add references}

An alternative view on rationality of choice is \emph{adaptationism}
\citep[e.g.,][]{Anderson1991:Is-human-cognit,ChaterOaksford2000:The-Rational-An,HagenChater2012:Decision-Making}. \todo{more
  references} Adaptationism aims to explain choice behavior by appeal to evolutionary
considerations: DMs have acquired choice mechanisms that outperform other manners of choice
making in a rich and variable environment. A choice mechanism could be a set of distinct
heuristics (the DM's \emph{adaptive toolbox}) that have little in common
\citep[e.g.,][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to relate to the philosophical debate about the nature of rational choice, we here suggest to
think of a choice mechanism as a map from choice situations to action choices that contains an
explicit level of subjective representation of a decision situation (see
Figure~\ref{fig:ChoiceMechanism}).

\begin{figure}
  \centering
  \begin{tikzpicture}[node distance = 2.5cm]

  \begin{scope}

    \node[] (environment) {};

    \node[rounded corners, draw, rectangle, minimum height = 0.75cm, text width=6cm, text
    centered, below of = environment, node distance = 2cm] (situation)
    {environment produces random \\ \textbf{decision situation} $d$};

    \node[rounded corners, draw, rectangle, minimum height = 0.75cm, text width=6cm, text
    centered, below of = situation] (representation) 
    {\textbf{subjective representation} $s$ of $d$ \\ \textcolor{gray}{[preference, beliefs, \dots]}};

    \node[rounded corners, draw, rectangle, minimum height = 0.75cm, text width=6cm, text centered, below of = representation] (action)
    {\textbf{action choice} $a$ based on $s$};

    \node[rounded corners, draw, rectangle, minimum height = 0.75cm, text width=6cm, text
    centered, below of = action] (payoff) {\textbf{payoff} for $a$ in $d$};

    \draw[->, thick]
    ([xshift=-1.75cm,yshift=-0.5cm]situation.center)--([xshift=-1.75cm,yshift=0.55cm]representation.center)
    node[midway, right] {subj.~representation scheme $\gamma$};

    \draw[->, thick]
    ([xshift=-1.75cm,yshift=-0.55cm]representation.center)--([xshift=-1.75cm,yshift=0.5cm]action.center)
    node[midway, right] {action selection function $\kappa$};

    \draw[->, thick]
    ([xshift=-1.75cm,yshift=-0.55cm]action.center)--([xshift=-1.75cm,yshift=0.5cm]payoff.center)
    node[midway, right] {};

    \node[minimum height = 0.75cm, text width=6cm, text
    centered, right of = representation, xshift = 2cm, rotate = 90] (mechanism) {choice mechanism};

    \begin{pgfonlayer}{background}
             \node [rounded corners, dashed,
       very thick,draw=black!40,fit={($(situation.south)+(0,-10pt)$) 
                                     ($(mechanism.west)+(+5pt,+42pt)$) 
                                     % ($(action.north)+(0,+10pt)$) 
                                     ($(representation.west)+(-7pt,0)$)
                                   }] (box) {};
    \end{pgfonlayer}

  \end{scope}

\end{tikzpicture}
  \caption{bla bla bla}
  \label{fig:ChoiceMechanism}
\end{figure}

 To make an adaptationist argument precise, we need
some

 When it comes to preferences, this is clearly
related to Alchian and Friedman's argument that firms \emph{should} optimize that which makes
them ``live,'' so to speak. For biological agents, the argument would be that subjective
preferences \emph{should} stay as close as possible to whatever determines objective fitness.
